---
title: "R for Reproducible Scientific Analysis"
output:
  html_document: default
  html_notebook: default
date: "18 January 2017"
---

# Creating Publication-Quality Graphics

> **Questions**
>
> - How can I create publication-quality graphics in R?
>
> **Objectives**
>
> - To be able to use `ggplot2` to generate publication quality graphics.
>
> - To understand the basic grammar of graphics, including the aesthetics and geometry layers, adding statistics, transforming scales, and coloring or panelling by groups.

Plotting the data is one of the best ways to quickly explore it and generate hypotheses about various relationships between variables.

There are several plotting systems in R, but today we will focus on `ggplot2` which implements **grammar of graphics** - a coherent system for describing components that consititute visual representation of data. For more information regarding principles and thinking behind `ggplot2` graphic system, please refer to [Layered grammar of graphics](http://vita.had.co.nz/papers/layered-grammar.pdf) by Hadley Wickham (@hadleywickham). 

The advantage of `ggplot2` is that it allows R users to create publication quality graphics with just a few lines of code. `ggplot2` has a large user base and is constantly developed and extended by the community.

## Getting started

`ggplot2` is a core member of `tidyverse` family of packages. Installing and loading the package under the same name will load all of the packages we will need for this workshop. Lets get started!

> Add a new chunk to this R Markdown Notebook by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

```{r}
# install.packages("tidyverse")
library("tidyverse")
```

If above code produces an error "there is no package called ‘tidyverse’", uncomment (remove #) the line above and run `install.packages()`command before you load the library. You only need to install the package once, but you will have to reload it, using the `library()` command, every time you restart R. 

Today we will be working with the `gapminder` dataset, which is the excerpt from the [GAPMINDER](http://www.gapminder.org/data/) data. First let's install and load the package containing the extracted data.

```{r}
#install.packages("gapminder")
library("gapminder")
```

You can have a look at the content of the `gapminder` data frame by simply typing `gapminder` either in the R-chunk or in the console. Data frame is a rectangular collection of data, where variables are organized as columns and observations are listed as rows.

```{r}
gapminder
```


The dataset contains the following fields:

- *country*:      country name
- *continent*:    continent name
- *year*:         year of observation
- *lifeExp*:      life expectancy at birth
- *pop*:          total population
- *gdpPercap*:    per-capita GDP

More information about the package and the data is available in help. Just type `?gapminder` in console, located in the bottom panel of your RStudio, or type `gapminder` in the search field of the *Help* tab of the bottom-right RStudio panel. Whenever you are unsure about anything in R, it is a good idea to check out the help file using one of the two methods described above.

## Creating the first plot

> Here's a question that we would like to answer using `gapminder` data: Do people in rich countries live longer then people in poor countries? The answer may be quite intitive, but we will continue our investigation further: how does the relationship between GDP per capita and Life expectancy look like? Is this relationship linear? Non-linear? Are there exceptions to the general rule (outliers)?

To plot `gapminder`, run the following code in the R-chunk or in console. The following code will put `gdpPercap` on the x-axis and `lifeExp` on the y-axis:

```{r}
ggplot(data = gapminder) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp))
```

Note that we split the function into two lines. The plus sign indicates that the function is not over yet and that the next line should be interpreted as additional layer to the preceding `ggplot()` function. In other words, when writing a `ggplot()` function spanning several lines, the `+` sign goes at the end of the line, not in the beginning.

The plot shows positive non-linear relationship between GDP per capita and Life expectancy. 

> Does this graph confirm or disprove your initial hypothesis about the relationship between these variables?

Note that in order to create a plot using `ggplot2` system, you should start your command with `ggplot()` function. It creates an empty coordinate system and initializes the dataset to be used in the graph (which is supplied as a first argument into the `ggplot()` function). In order to create graphical representaion of the data, we can add one or more layers to our otherwise empty graph. Functions starting with the prefix `geom_` create visual representation of data. In this case we added scattered points, using `geom_point()` function. There are many `geoms` in `ggplot2`, some of which we will learn in this lesson.

`geom_` function create *mapping* of variables from the earlier defined dataset to certain aestetic elements of the graph, such as axis, shapes or colors. The first argument of any `geom_` function expects the user to specify these mappings, wrapped in the `aes()` (short for *aestetics*) function. In this case, we mapped `gdpPercap` and `lifeExp` variables from `gapminder` dataset to x- and y-axis, respectively (using `x` and `y` arguments of `aes()` function). 

Generally speaking, the template for vizualizing data in `ggplot2` can be summarized as follows:


```{r, eval=FALSE}
`ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))`
```

In the remainer of this lesson we will learn how to extend and complete this template using different elements to produce various visualizations. First, we will look closer at the `<MAPPINGS>` component.

> **Challenge 1.**
>
> - How did Life expectancy change over time? What do you observe? Note that many points are plotted on top of each other. This is called "overplotting". Try different `geom_` function called `geom_jitter`. It will spread the points apart a little bit using random noise.
>
> *Hint: the* `gapminder` *dataset has a column called `year`, which should appear on the x-axis.*
>
> - See if you can visualize Life expectancy by continent. Which continent tends to have higher life expectancy (notice the density of the points along the y-axis)? Lowest life expectancy? Which continent has highest spread in life expectancy values? How about lowest spread? 

## Aestetic mappings

What if we want to combine graphs from previous two challenges and show the relationship between three variables in the same graph. Turns out, we don't necessarily need to use third geometrical dimension, we can simply employ color.

The following graph maps `continent` variable from `gapminder` dataset to the `color` aestetic of the plot. Let's take a look:

```{r}
ggplot(data = gapminder) + 
  geom_jitter(mapping = aes(x = year, y = lifeExp, color=continent))
```

> **Challenge 2.**
> 
> - What will happen if you switch the mappings of `continent` and `year` in the previous example? Is the graph still useful? Why? What if you map `color` aestetic to `country`? What has changed? How is `year` different from `country`? What is the limitation of the `color` aestetic, when used to vizualize different types of data? 
> 
> - Can you add a little color to our intial graph of life expectancy by GDP per capita? Color the points by continent. There seem to be some outliers in this graph. Can you now spot which continent to these points belong to? How about using color gradient to illustrate change over time?
>
> *Hint: you may want to transform GDP per capita to logarithmic scale before plotting. Just wrap the name of the variable into the* `log()` *function*

There are other aestetics that can come handy. One of them is `size`. The idea is that we can vary the size of datapoints to illustrate another continuous variable, such as country population. Lets look at four dimensions at once! 

```{r}
ggplot(data = gapminder) + 
  geom_point(mapping = aes(x = log(gdpPercap), y = lifeExp, color=continent, size=pop))
```

There's one more useful aestetic property of the graph which is good for visualizing low-cardinality categorical variables (categorical variables with small number of unique values), called `shape`. The idea is that you can employ different shapes (other than circles) to plot the data.

> **Challenge 3.**
> 
> - Blow your mind by visualizing five(!) dimensions in the same graph. Modify the previous example mapping year to color and shape to continent. What can you say about those Asian outliers: are those belonging to small or large countries? Are they from earlier or later time periods? 
>

Combining too many aestetics in the same graph can make it quite busy. However, you can always make remove certain aestatic properties and use several graphs to highlight different aspects of data.

Until now, we explored different aestetic properties of a graph mapped to certain variables. What if you want to recolor or use certain shape to plot all datapoints? Well, that means that such color or shape will no longer be *mapped* to any data, so you need to supply it to `geom_` function as a separate argument (outside of the `mapping`). Here's our intial graph with all colors colored in blue.

```{r}
ggplot(data = gapminder) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp), color="blue")
```

Once more, observe that the color is now not mapped to any particular variable from the `gapminder` dataset and applies equally to all datapoints, therefore it is ourside the `mapping` argument and is not wrapped into `aes()` function. Note that unmapped colors are supplied as characters (in quotes), size is a number (size of point in mm) and shape is the ordinal index of the shape in R's internal vocabulary (where square is 0, circle is 1, triangle is 2 and small filled circle is 20). Explore different shapes by varying the shape number between 0..25 or refer to `ggplot2` documentation, called [vignettes] (http://docs.ggplot2.org/current/vignettes/ggplot2-specs.html), for details. This document can be also called from within R by calling `vignette("ggplot2-specs")`. 

## Geometrical objects

Next, we will consider different options for <GEOM_FUNCTION> component of our template `ggplot2` graph example. Using different `geom_` functions user can highlight different aspects of data. For example, we could connect individual datapoints belonging to the same country into a line and illustrate the development of life expectancy over time for each country separately using `geom_line()` function. 

Some `geom_` functions allow additional aestetics, such as aestetic `group` in the `geom_line()` function. This aestetic may not have any meaning in other geoms, but here it allows us to draw multiple lines, one per country. To keep the lines organized, we will color them by continent. 

```{r}
ggplot(data = gapminder) + 
  geom_line(mapping = aes(x = year, y = lifeExp, group=country, color=continent))
```

Note how life expectancy suddently drops for certain countries in for short period of time. We will learn how to zoom in to those tragic periods of history and investigate which countries experienced them later in this workshop.

Another useful geom function is `geom_boxplot()`. It adds a layer with the "box and whiskers" plot illustrating the distribution of values within categories. The following chart breaks down life expectancy by continent, where the box represents first and third quartile (the 25th and 75th percentiles), the middle bar signifies the median value and the whiskers extent to cover 95% confidence interval. Outliers (outside of the 95% confidence interval range) are shown separately.

```{r}
ggplot(data = gapminder) + 
  geom_boxplot(mapping = aes(x = continent, y = lifeExp))
```

Layers can be added on top of each other. In the following graph we will place the boxplots over jittered points to see the distribution of outliers more clearly. We can map two aestetic properties to the same variable. Here we will also use different color for each continent.

```{r}
ggplot(data = gapminder) + 
  geom_jitter(mapping = aes(x = continent, y = lifeExp, color=continent)) +
  geom_boxplot(mapping = aes(x = continent, y = lifeExp, color=continent))

```

Now, this was slighly inefficient due to duplication of code - we had to specify the same mappings for two layers. To avoid it, you can move common arguments of `geom_` functions to the main `ggplot()` function. In this case every layer will "inherit" the same arguments, specified in the "parent" function.

```{r}
ggplot(data = gapminder, mapping = aes(x = continent, y = lifeExp, color=continent)) + 
  geom_jitter() +
  geom_boxplot()

```

You can still add layer-specific mappings or other arguments by specifying them within individual geoms. We would recommend building each layer separately and then moving common arguments up to the "parent" function.

We can use linear models to highlight differences in dependency between GDP per capita and life expectency by continent. Notice that we added a separate argument to the `geom_smooth()` function to specify the type of model we want `ggplot2` to built using the data (linear model). The `geom_smooth()` function has also helpfully provided confidence intervals, indicating "goodness of fit" for each model (shaded gray area). For more information on statistical models, please refer to help (by typing `?geom_smooth`)

```{r}
ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp, color=continent)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm")
```

Notice, that we also used new visual property called `alpha` to increase transparency of the data points and make trend lines stand out. `alpha` property can also be used as a mapping aestetic, i.e. transparency can be made to vary depending on the value of certain variable.

In the graph above, each geom inherited all three mappings: x, y and color. If we want only single linear model to be built, we would need to limit the effect of `color` aestetic to only `geom_point()` function, by moving it from the "parent" function to the layer where we want it to apply. Note, though, that because we want the `color` to be still mapped to the `continent` variable, it needs to be wrapped into `aes()` function and supplied to `mapping` argument.

```{r}
ggplot(data = gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(mapping=aes(color=continent), alpha=0.5) +
  geom_smooth(method="lm")
```

As you can observe the x-axis label of our graph says `log(gdpPercap)`, which indicates that we are not really plotting the original data, but rather the output of `log()` function. The same effect (with slightly more aestetically pleasing x-axis label) can be achieved by specifying the x-axis scale transformation as a separate layer.

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, color=continent)) +
  geom_point() +
  geom_smooth() +
  scale_x_log10()
```

## Faceting

Multi-layered graphs employing several aestetics can look crowded. In order to avoid it, one can split the data into different graphs using panels of similar graphs. In `ggplot2` this method is called "faceting". Lets facet the graph above by continent and show the datapoints and the trend for each continent in a separate chart.

```{r}
ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  geom_smooth() +
  scale_x_log10() + 
  facet_wrap( ~ continent)
```

The `facet_wrap()` layer takes a “formula” as its argument, denoted by the tilde (~). This tells R to draw a panel for each unique value in the `continent` column of the `gapminder` dataset. Faceting is useful when number of panels is limited. Notice that here R places panels from left to right, "wrapping" those panels that do not fit in one row onto the new line. Learn about advanced faceting, including faceting over several variables using help on `?facet_grid()`.

Reiterating our previously proposed `ggplot2` template and adding what we learned until, now we can state:

```{r, eval=FALSE}
`ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>)) + 
  <FACET_FUNCTION>`
```

## Statistical transformations

How many countries are there in `gapminder` dataset on each continent? It is to common to aswer this sort of question with a bar chart. This dataset contains observations from several years, but since we are no longer interested in showing the dynamics, perhaps it is good idea to take a "snapshot" of the data, by limiting it to observations from one year, say year 2007. In `tidyverse` terminology we will "filter" the data to contain only observations from 2007. We will learn about filtering later in this workshop. The simplest way of achieving it is to wrap the data and the condition we want to apply into the `filter()` function. 

```{r}
filter(gapminder, year==2007)
```

Now, each country is listed only once and we can answer the intial question: "How many countries are there on each continent"? Note that not every country of the world is included into the `gapminder` dataset (for unabridged version of the data, please, refer to `gapminder_unfiltered` dataset in the same package). 

We will use `geom_bar()` function to build a bar chart. Remember that we are using the filtered version of the data.

```{r}
ggplot(filter(gapminder, year==2007)) + 
  geom_bar(mapping = aes(x=continent))
```

Note the label on y-axis. It seems like R did something behind the scenes and instead of plotting row values it calculated and plotted summary of the data, in this case, count. We've already observed functions that calculated and plotted summaries of the data: boxplots and smoothers. This would be also true for functions producing histograms and other frequency functions.

The algorithm that calculates the data summary behind the scenes is called a `stat`, short of statistical transformation. Each geom function has a default `stat` associated with it. For simplicity default statistical transformations are ommitted from the function call, but they are there and can be overridden when necessary. The same chart as above can be produced by naming the stat explicitly.

```{r}
ggplot(filter(gapminder, year==2007)) + 
  geom_bar(mapping = aes(x=continent), stat="count")
```

In some cases we may not want R to calculate summary statistics for us. In that case we may want the height of the bars to correspond to certain values found in the data. This requires overriding default `stat` and instead using the reserved value, called `identity` (which means "use as is"). Lets show population of each country in Oceania in 2007.

```{r}
ggplot(filter(gapminder, year==2007 & continent=="Oceania")) + 
  geom_bar(mapping = aes(x=country, y=pop), stat="identity") 
```

Stats and geoms are cornerstones of `ggplot2` visualization toolbox. Seek more details in help, or refer to online documentation at http://ggplot2.org/

## Coordinate systems

There are only two countries from `Oceania` included in the dataset. Imagine that bars would be too narrow and labels would be unreadable. One way of dealing with it is flipping the coordiante system, i.e. plotting the same data as horizontal bars. Let's try to show population of every Asian country in 2007.

```{r}
ggplot(filter(gapminder, year==2007 & continent=="Asia")) + 
  geom_bar(mapping = aes(x=country, y=pop), stat="identity") +
  coord_flip()
```

There are many function related to coordinate systems that allow, among other things, plotting in non-cartesian (e.g. polar and mercantor) coordinates and specifying manual limits for coordinate axis. 

## Scales

We have been dealing with non-linear nature of the `gdpPercap` variable by plotting its log, which makes it really difficult to read and interpret the values of GDP per capita for individual countries. Instead of transforming the values, we could transform the scale of x-axis.

```{r}
ggplot(data = gapminder) + 
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp, color=continent)) +
  scale_x_log10() +
  facet_wrap(~year)
```

Now the x-axis is measured in log10 units and the data, plotted on log10 scale looks more linear. Certain scale and coordinate functions may result in similar visual effects on the chart, but the way they interact with other aestetic elements may be quite different. Check out the online `ggplot2` [documentation](http://docs.ggplot2.org/) for more details and examples of using scale and coordinate transformations.

## Labeling the chart

Lastly we will learn how to label and annotate the chart using `labs` and `annotate` functions. 

```{r}
ggplot(data = gapminder) + 
  geom_point(mapping = aes(x = gdpPercap/10^3, y = lifeExp, size=pop/10^6, color=continent)) +
  scale_x_log10() +
  facet_wrap(~year) + 
  labs(title="Life Expectancy vs GDP per capita over time",
       subtitle="In the past 50 years, life expectancy has improved in most countries of the world",
       caption="Source: Gapminder foundation, https://www.gapminder.org/data/",
       x="GDP per capita, '000 USD",
       y="Life expectancy, years",
       color="Continent",
       size="Population, mln")
```

## BONUS: Animation

The graph produced in the previous section looks quite good, but it requires a reader to follow the time aspect of the data by tracing the changes across panels. This may be better illustrated by "animating" the time dimension of the data and playing the twelve charts in front of user one after another.

```{r, fig.show = "animate"}
library(gganimate)
p <- ggplot(data = gapminder) + 
  geom_point(mapping = aes(x = gdpPercap/10^3, y = lifeExp, size=pop/10^6, color=continent, frame=year)) +
  geom_text(mapping = aes(x = 50000, y = 30, frame=year), label=as.character(gapminder$year), size = 16, color="grey") +
  scale_x_log10() +
  labs(title="Life Expectancy vs GDP per capita over time",
       subtitle="In the past 50 years, life expectancy has improved in most countries of the world",
       caption="Source: Gapminder foundation, https://www.gapminder.org/data/",
       x="GDP per capita, '000 USD",
       y="Life expectancy, years",
       color="Continent",
       size="Population, mln") +
  theme_bw()

gganimate(p)

```


## Wrap-up

We conclude this lesson by reiterating our `ggplot2` data visualization template. 

```{r, eval=FALSE}
`ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>),
                  stat = <STAT>) +
  <SCALE_FUNCTION> +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION> + 
  <LABS>`
```

We learned about seven parameters of ggplot functions. However, it is very rare that all six of them need to specified in a given graphic or chart. Most of the time ggplot offers useful defaults for everything other than data, geoms and mappings.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).
