---
title: "R for Reproducable Scientific Analysis"
author: "18 January 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Whenever you want to iterate, `purrr`!

R is a very compact language, optimized for working with vectors. This may come as a surprise to those who approach R with some knowledge of other programming languages. R is build in such a way that operations on vectors are made easy

```{r}
seq_len(5)^2/(2:6)
```

In order to make it happen, someone has written a pretty efficient `for` loop that efficiently executes these operations

```{r, eval=FALSE}
`for (i in 1:n) {`
  output[[i]] <- f(input[[i]])
}`
```

What if input is more complex than simple single-type, one-dimensional vector, generally speaking, a list. It may be worthwhile to remember what constitures a list and how to extract elements from a list. See [Vectors](http://r4ds.had.co.nz/vectors.html) chapter from "R for Data Science" by Hadley Wickham.

We will import another version of gapminder dataset from Julia's package `vega` and see if we can learn something about manipulating lists and building iterating structures using this dataset. Lets download and import the data and look at the structure

```{r, include = FALSE}
#install.packages("RJSONIO")
library(tidyverse)
url3 <- "https://raw.githubusercontent.com/johnmyleswhite/Vega.jl/master/vega-datasets/gapminder.json"
download.file(url3, "data/gapminder.json")
gap_list <- RJSONIO::fromJSON("data/gapminder.json", simplify = F) 
str(gap_list, max.level = 3, list.len = 3)
```

So this was a data stored in JavaScript Storage Object (json), which after importing with `RJSONIO::fromJSON()` becomes a list. 

> In this lesson we will first learn how to:
> 
> - pick out the data from the list using the `purrr` package.
>
> - iterate over vectors and lists to perform loop operations
> 
> - create and handle nested data frames using list-columns

# Handling lists

The ``purrr` package is the youngest member of `tidyverse` suite of packages. It is meant to be a more efficient implementation of the `apply` family of operations with stricter type-stability and some helpful shortcuts.

The central function of `purrr` package is `map`. It is made for applying a certain function to each element of a list (ref. `lapply()` in base R). Here's a useful template to remember:

```{r, eval=FALSE}
#FOR  EACH ELEMENT OF .X DO .F

` map(.x, .f, ...) `

# apply .f
# to every element of .x
# and return results _like_so

# basic map() usage
`map_OUT_TYPE(YOUR_LIST, YOUR_FUNCTION)`
```

There are many flavors of the `map()` function for each output type: `map_chr()`, `map_dbl()`, `map_int()`, etc. The default function `map()` without subscript always returns a list and in this sense equivalent in meaning to `map_list()`. There's a special function `walk()` that returns nothing.

The first attribute of the function is, conveniently, the data. This is needed for the pipe operators to work and, when included in the piped operation, the first element gets omitted.

The second attribute is a function. The simplest type of function is the subsetting, i.e. the famous `[` operator. Lets look around the data and try to extract the country list

```{r}
gap_list[1]
gap_list[[1]]
gap_list[[1]][1]

#extract country
map(gap_list, "country") %>% str(list.len = 3)
gap_list %>% map(~nchar(.x$country)) %>% str(list.len = 3)
```

We will now look at alternative outputs and alternative ways of specifying a function.

```{r}
gap_list %>% map_int(~nchar(.x$country)) %>% str()
gap_list %>% map_lgl(~(.x$life_expect<35)) %>% str()
gap_list %>% map_dbl(~log(.x$pop)) %>% str()
```

These functions hold no surprises. They return either vector of size .x or error.

Function was be also specified differently. Here's the case for selection/subsetting.

```{r}
## SUBSETTING BY NAME
gap_list %>% map_dbl("cluster") %>% str()
# equivalent to
gap_list %>% map_dbl(~.x[["cluster"]]) %>% str()

## SUBSETTING BY POSITION
gap_list %>% map_dbl(6) %>% str()
# equivalent to
gap_list %>% map_dbl(~.x[[6]]) %>% str()

```

Whenever we want to apply the same function to every element, we can also use some shortcuts

```{r}
gap_list %>% map_dbl(length) %>% str()
# equivalent to
gap_list %>% map_dbl(~length(.x)) %>% str()
```

`purr` can turn lists into tibbles

```{r}
gap_tbl <- tibble(
  year    =  gap_list %>% map_dbl("year"),
  country =  gap_list %>% map_chr("country"),
  cluster =  gap_list %>% map_dbl("cluster"),
  pop     =  gap_list %>% map_dbl("pop"),
  lifeExp =  gap_list %>% map_dbl("life_expect"),
  fertility  =  gap_list %>% map_dbl("fertility"))
gap_tbl

# alternative way
gap_tbl <- gap_list %>% map_df(`[`)
gap_tbl
```

finally, some of the columns of the dataset can be nested. Lets use use `tidyr` to summarize data per cluster.

```{r}
gap_nested_cluster <- gap_tbl %>% 
  group_by(cluster, year) %>% 
  summarise(pop=sum(pop),
            med_lifeExp=median(life_expect),
            fertility=mean(fertility),
            n_countries=n()) %>%
  group_by(cluster) %>%
  nest()
gap_nested_cluster
gap_nested_cluster$data[1]
```

The data inside the nested tibble is stipulated per country per year. Let's say we want to aggregate it to the year level.

```{r}
gap_nested_cluster_all <- gap_nested_cluster %>%
  mutate(plots = map(data, ~ggplot(.x)+geom_point(mapping=aes(x=year, y=fertility)))
         ) 
dir.create(file.path("output"))
gap_nested_cluster_all %>% select(plots) %>% walk(print)
walk2(gap_nested_cluster_all$cluster, gap_nested_cluster_all$plots,
      ~ggsave(paste0("output/cluster_", .x, ".png"),.y))

```



# Using `dplyr`, `ggplot2` and `purrr` to explore a data set

It is intruiging to investigate data sets like `gapminder` for patterns. In this case , we're going to focus on just three variables to answer the question "How does life expectancy (`lifeExp`) change over time (`year`) for each country (`country`)?". A good place to start is with a plot:

```{r}
gapminder <- readr::read_csv(file = "data/gapminder-FiveYearData.csv")

gapminder %>% 
  ggplot() + 
  geom_line(mapping = aes(x = year, y = lifeExp, group = country, color = continent))
```

There seems to be an overall linear trend. Let's have a more detailed look, dividing the data set by continent:

```{r}
gapminder %>% 
  ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, color = country)) +
    geom_smooth(mapping = aes(x = year, y = lifeExp), method = "lm", color = "black") +
    facet_wrap( ~ continent) +
    theme(legend.position = "none")
```

There seems to be a good handful of countries with a pronounced irregularity. We will now produce a linear model for each country. But instead of repeating an action for each variable, we want to repeat an action for each country, a subset of rows. To do that, we need a new data structure: the nested data frame. To create a nested data frame we start with a grouped data frame, and "nest"" it:

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()
by_country
```

Now that we have our nested data frame, we're in a good position to fit some models. We will use a model-fitting function which we want to apply it to every data frame. 

```{r}
build_country_model <- function(x) {
  lm(formula = lifeExp ~ year, data = x)
}
```

The data frames are in a list, so we can use `map()` to apply `country_model` to each element. Instead of creating a new object, we're going to create a new variable in the `by_country` data frame. That's a job for `mutate()`.

```{r}
by_country <- by_country %>% 
  mutate(model = purrr::map(data, build_country_model))
by_country
```

We have now created a linear model for every country! To take a better look at the model quality metrics, *e.g.* the R^2^, we will use the function `glance()` from the `broom` package, which is already loaded as part of `tidyverse`. If we apply it to a model, we get a data frame with a single row. We can use `mutate()` and `unnest()` to create a data frame with a row for each country. To suppress list columns (for data frames and models), we use `.drop = TRUE`.

```{r}
glance_df <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE)
glance_df
```

With this data frame in hand, we can start to look for models that don’t fit well:

```{r}
glance_df %>% 
  arrange(r.squared)
```

The worst models all appear to be in Africa. Let's double check that with a plot. Here we have a relatively small number of observations and a discrete variable, so `geom_jitter()` is effective:

```{r}
glance_df %>% 
  ggplot() + 
    geom_jitter(mapping = aes(x = continent, y = r.squared, color = continent), width = 0.3)
```

We could pull out the countries with particularly bad R^2^ and plot the data:

```{r}
bad_fit <- glance_df %>% 
  filter(r.squared < 0.5)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, colour = country))
```

We see two main effects here: the tragedies of the HIV/AIDS epidemic and the Rwandan genocide.
