---
title: "R for Reproducible Scientific Analysis"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  html_notebook: default
date: "25 September 2018"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages("jsonlite")
# install.packages("remotes")
# remotes::install_github("dmi3kno/googlesheets")
library(tidyverse)
#gapminder <- gapminder::gapminder
library(gapminder)

```
## Introduction to lists

We would like to go beyond the few columns provided in the `gapminder` dataset and seek additional data that could allow us do other types of analyses and/or look at the data in new dimensions. 

[GAPMINDER website](https://www.gapminder.org/data/) contains quite a few files, which can be downloaded in xls format. We will download the list of available files and use it to learn how to extract pieces of information from nested lists.

Lists are common in R and can be stored in various binary formats, such as (.Rds, .Rdata, .Rda). These file formats are object-agnostic and can contain anything, since objects are stored in serialized form. It is also possible to store a list as text in "JavaScript Object Notation" (json) file, a cross-platform format for storing stuctured information, which comes from JavaScript. 

Here's a list of files hosted on GAPMINDER website in json format.
```{r}
#txt <- read_lines("https://www.gapminder.org/wp-content/themes/gapminder/cronJobs/json.js")
#lst <- jsonlite::fromJSON(str_c("[", str_extract(txt, "\\{.+\\}"),"]"), simplifyVector = FALSE)[[1]]
#write_lines(jsonlite::toJSON(lst), "gapminder_data_table.json")

#gap_lst <- jsonlite::fromJSON("https://raw.githubusercontent.com/dmi3kno/SWC-tidyverse/master/gapminder_data_table.json")

gap_lst <- jsonlite::fromJSON("https://bit.ly/gapminderlist")

typeof(gap_lst)

```

As we already know, list is a special type of vector.  Lists have length. 

```{r}
length(gap_lst)
```

There are several ways of subsetting a list. Some of these methods were covered in pre-read. List nodes may be named. Names are optional, but sometimes they make navigation easier. 

Print method of lists is not most user-friendly, as it tries to output the whole list to the console. Throughout this lesson we will use `head()` and `tail()` throughout to conserve screen space. Remember to remove these functions when you want the whole list (e.g. when assigning output to a variable).

```{r, eval=FALSE}
gap_lst %>% head() %>% 
  str()

gap_lst[3] %>% str()
gap_lst[[3]] %>% str()
```

List elements may be named. The names of the top-level elements of `gap_lst` are a googlesheet codes. We will use `$` notation, which is a special notation for named lists (you can also use it to create new elements in the list). 

```{r, eval=FALSE}
gap_lst$phAwcNAVuyj0TAlJeCEzcGQ
gap_lst$phAwcNAVuyj0TAlJeCEzcGQ$indicatorName

```

Most important/useful feature of lists, is that they can contain other vectors (lists are called "recurring vector"), e.g. list has branches that contain similar entities. In our case each branch in the main list contains a list of 5 elements (5 "leaves"). If we want to extract one particular leaf from every branch, we need to "iterate" over the branches, visiting every one of then and selecting an element of interest.

```{r}
gap_lst %>% 
  map_chr("indicatorName") %>% 
  head()
```

The function `map` is equivalent to for-loop. Normally `map` expects a list as first argument and a function as a second argument. One of the cool features of `map` is that we can also we pass string as second argument. In this case it acts as "accessor" - required elements get extracted from the list. Just remember that `map` can be used to extract an element of the iterated sequence. It visits each element of original vector (each child) and looks for nodes among grand-children. It can look by name or by index.

The resulting type is controled by the suffix after the function name. For example, if we want to receive character vector, we can pass `map_chr`. For integers it would be `map_int`, for real numbers `map_dbl`. Most generic type of function is `map` without suffix, which always returns a list. 

## Assignment 1 {.tabset .tabset-fade .tabset-pills}

### Question
>
> - Extract all googlesheet IDs. What function can help you with this task?
> - Extract a vector of data providers. What are the most common data providers? *Tip: You may find `table` function useful*

### Answer

```{r}
names(gap_lst) %>% head()

gap_lst %>% map_chr(4) %>% table() %>% sort() %>% tail()
gap_lst %>% map_chr("dataprovider") %>% table() %>% sort() %>% tail()
```

## Using `map` for iteration

Each element in original `gap_lst` was also a list. It stores various information about the googlesheet files. Think about it as a list of files with various information about each file. Since each of the elements is a list, it should have its own length. How long is each of these lists? Well, we could visit each "branch" and call `length()` function on it.

```{r}
gap_lst[[1]] %>% length()
gap_lst[[2]] %>% length()
# etc.
```

This is a perfect place to apply a loop. But as [Hadley Wikham argues](https://www.youtube.com/watch?v=GyNqlOjhPCQ) there's much more expressive, concise and succint way of implementing iteration - functional programming. `purrr` offers `map` family of functions that encapsulate and abstract for-loops.

We can iterate over elements of `gap_lst` and apply the same `length()` function to each element of the list.

```{r}
map_int(gap_lst, length) %>% head()
```

The suffix after the name of the function specifies expected return type. In this case we say that we want to iterate over the elements of the `gap_lst` and apply function `length` to each of them. The result, we expect to be an integer vector (since length is measured in whole numbers). This ensures type-stability. There are similar variants of `map` function for returning other base types: logical (`map_lgl`), real(`map_dbl`) and character(`map_chr`). Most generic function of all is `map` itself, it always returns a list.


## Transposing lists

We can transpose the list using the function `transpose`. You can think of this operation as "filpping" the relationship betwen parents and children: children become parents and vice versa. Let see how long is the list now. `transpose` is very often followed by another useful function `simplify`, which basically tries to convert lists to vectors, where possible. `simplify_all` will do it for every element of a list.

```{r}
gap_lst_trans <- transpose(gap_lst) %>% simplify_all()
# gap_lst_trans %>% View()
length(gap_lst_trans)
```

As you might have noticed, we are dealing with "named" vectors. These are special type of vectors where each element carries a name. Sometimes it is useful, but very often it just makes print out really noisy. In our case names are repeated in each vector as result of `transpose` opertation we did on the original list. 

### Lifehack 1: Naming things

> You can name things by composing a vector of names corresponding to the vector of things you want to name and then declare former to be the `names` of the latter. Be careful! The names will stick!
>
> You can undo your spell by calling `unname`. Remember that names are sticky, so you will have to override (overwrite). Be careful: certain things shall not be unnamed (such as columns in a dataframe).

```{r}
heights <- c(182, 119, 120)
names(heights) <- c("Dmytro", "Ian", "Mark")

heights

unname(heights)
heights

heights <- unname(heights)
heights
```

Lets try to extract googlesheet IDs and add them as separate element to our list. We can use `$` notation here.

Note that each of the vector in the list is now of equal length. This condition is enough to call our list a data frame.

> Recall that dataframe is a list of vectors of equal lengths.

We will remove the names from the vectors using unname function and return the result in the tibble format.

```{r}
# named vector
gap_lst_trans$indicatorName %>% head
names(gap_lst_trans$indicatorName) %>% head
gap_lst_trans$googlesheet_id <- names(gap_lst_trans$indicatorName)
gap_lst_df <- map_df(gap_lst_trans, unname)
gap_lst_df
```

## Applying functions to lists

Transposed list is a little more useable. Lets practice some more with applying functions to lists. We can create, so called "anonimous functions" but defining a function body right there in the .f argument of the `map` function. This way we save us some typing and effort of naming. This makes sense if we are pretty sure we will not have to reuse the same function ever again. Here's a function for calculating average length of the words in each element of transposed list. 

```{r}
nchar("You")

nchar(c("You", "me"))

c("You", "me", "us", "them") %>% nchar() %>% mean()


avg_char_length <- function(x){
  mean(nchar(x))
}

c("You", "me", "us", "them") %>% avg_char_length()

gap_lst_trans %>% map_dbl(avg_char_length)

gap_lst_trans %>% map_dbl(function(x) mean(nchar(x)))
gap_lst_trans %>% map_dbl(~ mean(nchar(.x)))
```

First option is to write a function exactly how one would write a regular function, except without naming it. It is a good idea if you dont intend to reuse it elsewhere and therefore would like to avoid naming it (one of the two hardest things in programming).

Second option utilizes a shortcut notation involving a `~` sign. It also has "reserved" argument name(s) `.x`, `.y`. and `.z` for first, second and third argument, respectively. Having these shortcuts allow us to indicate where we want the iterated data to "land" in our formula. This is particularly useful in case of multiple arguments utilized in `map2_` functions, which allow iteration over two vectors simultaneously.

## Assignment 2 {.tabset .tabset-fade .tabset-pills}

### Question
> Remember that tibbles are exalted lists. Exploit that property to anser the following questions:
>
> - How many columns in `gapminder` dataset are numeric?
> - Count "" values in each column of `gap_lst_df`. *Tip: string '' has zero characters*

### Answer

```{r}

gapminder %>% 
  map_lgl(is.numeric) %>% 
  sum()

gap_lst_df %>% 
  map_int(~sum(.x==""))

gap_lst_df %>% 
  map_int(~sum(nchar(.x)==0))
```

## Writing your own function

Functional programming starts with creating a function. Function has a name, a set of arguments and a body. Lets implement a simple function for checking if the number is even.

```{r}
is_even <- function(x){
  x %% 2 == 0 
}

is_even(26)
is_even(31)
```

This pretty simple function has only one argument that we decided to call `x`. It is a good idea to make argument names a little bit more expressive, unless a function is pretty generic, in which case `x` is a reasonably common and well-accepted name for an argument (in our case a character input). Note, that because we used vectorized operation inside our function, this function will also work on vectors.

```{r}
is_even(1:5)
```

Now that we are cetain our function is vectorized, we can use it to create new columns in a data frame (tibble). Remember that in order to create new column we can use the function `mutate`

```{r}
gapminder %>% 
  mutate(pop_even=is_even(pop))
```

## Assignment 3 {.tabset .tabset-fade .tabset-pills}

### Question
>
> - Write a function for calculating leap years. 
>
> Wikipedia: Every year that is exactly divisible by four is a leap year (`x%%4==0`), except for years that are exactly divisible by 100 (`x%%100!=0`), but these centurial years are leap years if they are exactly divisible by 400 (`x%%400==0`). For example, the years 1700, 1800, and 1900 were not leap years, but the years 1600 and 2000 were
> 
> - Add a column to gapminder indicating if the year is a leap year

### Answer

```{r}
is_leap_year <- function(x){
  x%%4==0 & (x%%100!=0 | x%%400==0)
}

is_leap_year(1700)
is_leap_year(2018)


gapminder %>% 
  mutate(leap_year = is_leap_year(year))
```


## Map-reduce

Whenver you are working with lists, you might want to apply a function to "reduce" or "collapse" a set of values to a single value using a function (say you got a list of numbers and you would like to add them up). Or you have a list of logicals and you would like to check if at least one of them is `TRUE`. In these cases you may want to use `reduce`. Reduce is fundamentally different from vectorized functions that we all are familiar with by now in how it performs desired operation on the data. It starts from element #1 and performs the desired operation with element #2. The result is processed together with element #3 and so on. Effectively, it is performing a "chain" operation, where result of previous iteration is input into the next one. 

```{r}
c(FALSE, TRUE, FALSE, FALSE) %>% any()

c(FALSE, TRUE, FALSE, FALSE) %>% all()

map(1:5, is_even) %>% 
  reduce(any)

map(1:5, ~.x*2) %>% 
  reduce(`+`)

```

## Side note on googlesheets

Gapminder data is published as collection of Google spreadsheets. There's a package for handling this type of data, called `googlesheets`. Have a look at the [readme](https://github.com/jennybc/googlesheets) and  [vignette](https://rawgit.com/jennybc/googlesheets/master/vignettes/basic-usage.html) of this awesome package for details and usage examples. 


Lets practice by importing a sheet containing more information about Gapminder countries  using this package. This sheet is linked on [Gapminder website](https://www.gapminder.org/data/geo/), found under Gapminder-Data-Geography. You may need to authenticate access to google drive, since this spreadsheet is private. You can also download this spreadsheet from the website in `xlsx` format and import it with `readxl::read_excel()` command.

```{r}
library(googlesheets)
#gs_auth() 

geo_sheet <- gs_url("https://docs.google.com/spreadsheets/d/1qHalit8sXC0R8oVXibc2wa2gY7bkwGzOybEMTWp-08o/")

x<- gs_key("1qHalit8sXC0R8oVXibc2wa2gY7bkwGzOybEMTWp-08o")

geo_sheet %>% 
  gs_read(ws="list-of-countries-etc")

```
## More functions

## Assignment 4 {.tabset .tabset-fade .tabset-pills}

### Question
>
> - Write a function for downloading Data from googlesheets and writing to csv files. The function will take a google key, retrieve sheet information using `gs_key`, read the sheet named "Data" with `gs_read` and write the content into a csv file named like `paste0(x, ".csv")`, where `x` is your google sheet key.

### Answer

```{r}
save_gs <- function(x){
  key <- gs_key(x, lookup = FALSE)
  sheet <- gs_read(key, ws="Data")
  write_csv(sheet, paste0(x, ".csv"))
}

fert_key <- "1oq3r8W7ajenKFgoAYoOf2MXeTWWNPpudR-Fo5m2-o30"
save_gs(fert_key)

```

## Scaling up the knowledge

> Use any of the known to you methods to extract google sheet ids for the following files: 
> 
> - Children per woman (total fertility) -  (category = "" and subcategory = "")
> - Infant mortality (rate per 1,000 births)	-  (category = "Health" and subcategory=""),
> - CO2 emissions (tonnes per person) (category = "" and subcategory = ""), and
> - Inequality index (Gini)	- (category = "Economy" and subcategory= "Poverty & Inequality")
> 
> Combine four googlesheet keys into a singe vector and create another vector with names of the variables you want the data from these sheets to end up in (e.g. "fert" or "mort")

Below is the example code for combining two files and joining them to `gapminder` dataset. Extend this code to be able to join any number of files from gapminder.org website. You can use the `gap_lst_df` we created earlier.

```{r, error=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(gapminder)
library(googlesheets)

keys <- c("phAwcNAVuyj0TAlJeCEzcGQ","phAwcNAVuyj0NpF2PTov2Cw","phAwcNAVuyj1gkNuUEXOGag","pyj6tScZqmEcjeKHnZq6RIg")
vars <- c("fert", "mort", "emissions", "gini")

importSheet <- function(key, var_name){
  sheet <- gs_key_old(key, lookup = FALSE)
  sheet_df <- gs_read(sheet)
  names(sheet_df)[1] <- "country"
  sheet_df_long <- sheet_df %>% 
    gather(key="year", value="value", -country) %>% 
    mutate(year=as.numeric(year), value=as.numeric(value))
  names(sheet_df_long)[3] <- var_name
  sheet_df_long
}

gapminder_plus <- map2(keys,vars, importSheet) %>% 
  reduce(full_join, by=c("country", "year")) %>% 
  right_join(gapminder, by=c("country", "year"))

write_csv(gapminder_plus, "data/gapminder_plus.csv")
```


