---
title: "R for Reproducable Scientific Analysis"
author: "18 January 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
keys <- tribble( ~key, ~var_name,
  "1oq3r8W7ajenKFgoAYoOf2MXeTWWNPpudR-Fo5m2-o30", "fert",
  "1OHMMuHbSFKDolNHXsmgHYlkjSKfAZyyY1P-ddMu_Fz0", "infantMort")

library(googlesheets)

importSheet <- function(key, var_name){
  sheet <- gs_key(key, lookup = FALSE)
  sheet_df <- gs_read(sheet)
  names(sheet_df)[1] <- "country"
  sheet_df_long <- sheet_df %>% 
    gather(key="year", value="value", -country) %>% 
    mutate(year=as.numeric(year), value=as.numeric(value))
  names(sheet_df_long)[3] <- var_name
  return(sheet_df_long)
}

gapminder_plus <- keys %>% split(.$key)  %>%  
  map(~importSheet(.$key, .$var_name)) %>% 
  reduce(full_join, by=c("country", "year")) %>% 
  right_join(gapminder, by=c("country", "year"))

summary(gapminder_plus)
```

## Whenever you want to iterate, `purrr`!

R is a very compact language, optimized for working with vectors. This may come as a surprise to those who approach R with some knowledge of other programming languages. R is build in such a way that operations on vectors are made easy

```{r}
seq_len(5)^2/(2:6)
```

In order to make it happen, someone has written a pretty efficient `for` loop that efficiently executes these operations

```{r, eval=FALSE}
`for (i in 1:n) {`
  output[[i]] <- f(input[[i]])
}`
```

Simplest example of iteration is what we have already discovered in the `dplyr` lesson - using summary functions within `group_by`. The disadvantage of this approach is that you need to specify all of the designed transformations upfront, otherwise richness of data will be gone, once the data is collapsed into the final summary statistics.

```{r}
## conventional summary
gapminder_plus %>% group_by(continent) %>% 
  summarise(min_lifeExp=min(lifeExp))

## two summary functions at the same time
gapminder_plus %>% group_by(continent) %>% 
  summarise_each(funs(min, max), lifeExp)

```

How can we delay the "collapsing" of the rich yearly data into the summary statistics without cluttering up the data frame? How can we summarize the data that includes more complex objects than single values (models, graphs, etc)?

# Using `dplyr`, `ggplot2` and `purrr` to explore a data set

It is intruiging to investigate data sets like `gapminder` for patterns. In this case , we're going to focus on just three variables to answer the question "How does life expectancy (`lifeExp`) change over time (`year`) for each country (`country`)?". A good place to start is with a plot:
_
```{r}

gapminder_plus %>% 
  ggplot() + 
  geom_line(mapping = aes(x = year, y = lifeExp, group = country, color = continent))
```

There seems to be an overall linear trend. Let's have a more detailed look, dividing the data set by continent:

```{r}
gapminder %>% 
  ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, color = country)) +
    geom_smooth(mapping = aes(x = year, y = lifeExp), method = "lm", color = "black") +
    facet_wrap( ~ continent) +
    theme(legend.position = "none")
```

There seems to be a good handful of countries with a pronounced irregularity. We will now produce a linear model for each country. But instead of repeating an action for each variable, we want to repeat an action for each country, a group of rows. Using "tidyverse" family of packages, it is possible to package the data belonging to each country and place it into a separate column as a "nested" data frame:

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()
by_country
```

What can we notice from this format. First of all our data frame have shrunk to only 142 obserations - one for each country. The identifiers (country and continent) are still present in the dataset. There's a new data variable - by the type "tibble". it is possible to extract the nested elements of this data structure.

```{r}
by_country[["data"]][1] 
```


Now that we have our nested data frame, we're in a good position to fit some models. We will use a model-fitting function which we want to apply it to every data frame. 

```{r}
build_country_model <- function(x) {
  lm(formula = lifeExp ~ year, data = x)
}
```

The data frames are in a list, so we can use `map()` to apply `country_model` to each element. Instead of creating a new object, we're going to create a new variable in the `by_country` data frame. That's a job for `mutate()`.

```{r}
by_country <- by_country %>% 
  mutate(model = purrr::map(data, build_country_model))
by_country
```

We have now created a linear model for every country! To take a better look at the model quality metrics, *e.g.* the R^2^, we will use the function `glance()` from the `broom` package, which is already loaded as part of `tidyverse`. If we apply it to a model, we get a data frame with a single row. We can use `mutate()` and `unnest()` to create a data frame with a row for each country. To suppress list columns (for data frames and models), we use `.drop = TRUE`.

```{r}
glance_df <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE)
glance_df
```

With this data frame in hand, we can start to look for models that don’t fit well:

```{r}
glance_df %>% 
  arrange(r.squared)
```

The worst models all appear to be in Africa. Let's double check that with a plot. Here we have a relatively small number of observations and a discrete variable, so `geom_jitter()` is effective:

```{r}
glance_df %>% 
  ggplot() + 
    geom_jitter(mapping = aes(x = continent, y = r.squared, color = continent), width = 0.3)
```

We could pull out the countries with particularly bad R^2^ and plot the data:

```{r}
bad_fit <- glance_df %>% 
  filter(r.squared < 0.5)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, colour = country))
```

We see two main effects here: the tragedies of the HIV/AIDS epidemic and the Rwandan genocide.
