---
title: "R for Reproducable Scientific Analysis"
author: "5 April 2017"
output:
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library("tidyverse")
gapminder <- read_csv(file = "../Data/gapminder-FiveYearData.csv")
```

# Dataframe Manipulation with `dplyr`

Manipulation of dataframes means many things to many researchers, we often select certain observations (rows) or variables (columns), we often group the data by a certain variable(s), or we even calculate summary statistics. Subsetting data sets can be challenging in base R, because there is a fair bit of repetition. Repeating yourself will cost you time, both now and later, and potentially introduce some nasty bugs.

# Key Points

* Use the `dplyr` package to manipulate dataframes.
* Use `select()` to choose variables from a dataframe.
* Use `filter()` to choose data based on values.
* Use `group_by()` and `summarize()` to work with subsets of data.
* Use `mutate()` to create new variables.

# The `dplyr` package

The [`dplyr`](https://cran.r-project.org/web/packages/dplyr/dplyr.pdf) package provides a number of very useful functions for manipulating dataframes in a way that will reduce the above repetition, reduce the probability of making errors, and probably even save you some typing. As an added bonus, you might even find the `dplyr` grammar easier to read.

`dplyr` is part of the `tidyverse` package collection, so it should already be loaded. If you've started a new session, load the `tidyverse package:

```{r}
library("tidyverse")
```

# The pipe `%>%`

We often want to string together a series of `dplyr` functions. This is achieved using `dplyr`'s pipe operator, `%>%`. This takes the value on the left, and passes it as the first argument to the function call on the right. 

`%>%` isn’t limited to `dplyr` functions. It's an alternative way of writing any `R` code:

```{r}
rep("This is an example", times = 3)
"This is an example" %>% rep(times = 3)
```

The shortcut to insert the pipe operator is `Ctrl`+`Shift`+`M` for Windows/Linux, and `Cmd`+`Shift`+`M` for Mac.

# Using `select()`

If, for example, we wanted to move forward with only a few of the variables in our dataframe we could use the `select()` function. This will keep only the variables you select.

```{r}
year_country_gdp <- select(gapminder, year, country, gdpPercap)
year_country_gdp
```

## <span style="color:red">Show Figure 1</span>

If we open up `year_country_gdp` we'll see that it only contains the `year`, `country` and `gdpPercap`. Above we used 'normal' grammar, but the strengths of `dplyr` lie in combining several functions using pipes. Since the pipes grammar is unlike anything we've seen in `R` before, let's repeat what we've done above using pipes.

```{r}
year_country_gdp <- gapminder %>% 
  select(year, country, gdpPercap)
year_country_gdp
```

To help you understand why we wrote that in that way, let's walk through it step by step. First we summon the `gapminder` dataframe and pass it on, using the pipe symbol (`%>%`), to the next step, which is the `select()` function. In this case we don't specify which data object we use in the `select()` function since in gets that from the previous pipe. Fun fact: There is a good chance you have encountered pipes before in the shell. In `R`, a pipe symbol is `%>%` while in the shell it is `|` but the concept is the same!

# Using `filter()`

The `filter` function is a intuitive way of subsetting data. For example, we can look at the population per continent in 2002:

```{r}
gapminder %>% 
  filter(year == 2002) %>% 
  ggplot(mapping = aes(x = continent, y = pop)) +
  geom_boxplot()
```

If we now wanted to move forward with the above, but only with European countries, we can combine `select()` and `filter()`:

```{r}
year_country_gdp_euro <- gapminder %>% 
  filter(continent == "Europe") %>%
  select(year, country, gdpPercap)
year_country_gdp_euro
```

## <span style="color:red">Challenge 1</span>

Write a single command (which can span multiple lines and includes pipes) that will produce a dataframe that has the Norwegian values for `gdpPercap`, `lifeExp` and `year`, but not for other countries How many rows does your dataframe have and why?

```{r}
country_lifeExp_Norway <- gapminder %>%
  filter(country == "Norway") %>%
  select(year, lifeExp, gdpPercap)
country_lifeExp_Norway
```

As with last time, first we pass the gapminder dataframe to the `filter()` function, then we pass the filtered version of the gapminder dataframe to the `select()` function. **Note**: The order of operations is very important in this case. If we used `select()` first, `filter()` would not be able to find the variable continent since we would have removed it in the previous step.

# Using `group_by()`

Now, we were supposed to be reducing the error prone repetitiveness of what can be done with base `R`, but up to now we haven't done that since we would have to repeat the above for each continent. Instead of `filter()`, which will only pass observations that meet your criteria (in the above: `continent == "Europe"`), we can use `group_by()`, which will essentially use every unique criteria that you could have used in filter.

```{r}
gapminder
gapminder %>% 
  group_by(continent)
```

## <span style="color:red">Show Figure 2</span>

# Using `summarize()`

The above was a bit on the uneventful side because `group_by()` much more exciting in conjunction with `summarize()`. This will allow use to create new variable(s) by using functions that repeat for each of the continent-specific data frames. That is to say, using the `group_by()` function, we split our original dataframe into multiple pieces, then we can run functions (e.g. `mean()` or `sd()`) within `summarize()`.

```{r}
gapminder %>% 
  group_by(continent) %>%
  summarize(mean_gdpPercap = mean(gdpPercap))
```

We can also use a pipe to directly "feed" our processed data into `ggplot2`:

```{r}
gapminder %>% 
  group_by(continent) %>%
  summarize(mean_gdpPercap = mean(gdpPercap)) %>% 
  ggplot(mapping = aes(x = continent, y = mean_gdpPercap)) +
  geom_point()
```

## <span style="color:red">Show Figure 3</span>

That allowed us to calculate the mean `gdpPercap` for each continent, but it gets even better.

## <span style="color:red">Challenge 2</span>

Calculate the average life expectancy per country in Asia. Which has the longest average life expectancy and which has the shortest average life expectancy?

```{r}
gapminder %>% 
  filter(continent == "Asia") %>% 
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>% 
  filter(mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))
```

The function `group_by()` allows us to group by multiple variables. Let's group by year and continent.

```{r}
gapminder %>% 
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap))
```

That is already quite powerful, but it gets even better! You're not limited to defining one new variable in `summarize()`.

```{r}
gapminder %>% 
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap),
            sd_gdpPercap = sd(gdpPercap),
            mean_pop = mean(pop),
            sd_pop = sd(pop))
```

# Using `mutate()`

We can also create new variables prior to (or even after) summarizing information using `mutate()`.

```{r}
gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9)

gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9) %>%
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap),
            sd_gdpPercap = sd(gdpPercap),
            mean_pop = mean(pop),
            sd_pop = sd(pop),
            mean_gdp_billion = mean(gdp_billion),
            sd_gdp_billion = sd(gdp_billion))
```

## <span style="color:red">Challenge 3</span>

Calculate the average life expectancy and GDP in 1987 for each continent.

```{r}
gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9) %>%
  filter(year == 1987) %>%
  group_by(continent) %>%
  summarize(mean_lifeExp = mean(lifeExp),
            mean_gdp_billion = mean(gdp_billion))
```

# Using `dplyr` and `ggplot2` to visualize average global life expectancy

We will now demonstrate how to use functions from both `dplyr`and `ggplot2` to visualize the average life expectancy of countries that are part of the `gapminder` data set.

First, we need to calculate the average life expectancy for each country using the `summarize()` function:

```{r}
gapminder_country_summary <- gapminder %>% 
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp))
gapminder_country_summary
```

Next, we use a database for mapping coordinates (`install.packages("maps")`) to align with the average life expectancy data from above. We can directly use the output and pipe it into `ggplot()`! 

```{r}
map_data("world") %>% 
  rename(country = region) %>% 
  left_join(gapminder_country_summary, by = "country") %>% 
  ggplot() +
  geom_polygon(aes(x = long, y = lat, group = group, fill = mean_lifeExp)) +
  scale_fill_gradient(low = "blue", high = "red") +
  coord_equal()
```

Note that "grey" countries are not part of `gapminder` or appear under a different name and hence only appear as `NA`'s.

# Key Points

* Use the `dplyr` package to manipulate dataframes.
* Use `select()` to choose variables from a dataframe.
* Use `filter()` to choose data based on values.
* Use `group_by()` and `summarize()` to work with subsets of data.
* Use `mutate()` to create new variables.
