---
title: "R for Reproducible Scientific Analysis"
author: "18 January 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
gapminder <- gapminder::gapminder

keys <- tribble( ~key, ~var_name,
  "1oq3r8W7ajenKFgoAYoOf2MXeTWWNPpudR-Fo5m2-o30", "fert",
  "1OHMMuHbSFKDolNHXsmgHYlkjSKfAZyyY1P-ddMu_Fz0", "infantMort")

library(googlesheets)

importSheet <- function(key, var_name){
  sheet <- gs_key(key, lookup = FALSE)
  sheet_df <- gs_read(sheet)
  names(sheet_df)[1] <- "country"
  sheet_df_long <- sheet_df %>% 
    gather(key="year", value="value", -country) %>% 
    mutate(year=as.numeric(year), value=as.numeric(value))
  names(sheet_df_long)[3] <- var_name
  return(sheet_df_long)
}

gapminder_plus <- keys %>% split(.$key)  %>%  
  map(~importSheet(.$key, .$var_name)) %>% 
  reduce(full_join, by=c("country", "year")) %>% 
  right_join(gapminder, by=c("country", "year"))

write_csv(gapminder_plus, "data/gapminder_plus.csv")
```

## Whenever you want to iterate, `purrr`!

R is a very compact language, optimized for working with vectors. This may come as a surprise to those who approach R with some knowledge of other programming languages. R is build in such a way that operations on vectors are made easy

```{r}
seq_len(5)^2/(2:6)
```

In order to make it happen, someone has written a pretty efficient `for` loop that efficiently executes these operations

```{r, eval=FALSE}
`for (i in 1:n) {`
  output[[i]] <- f(input[[i]])
}`
```

Simplest example of iteration is what we have already discovered in the `dplyr` lesson - using summary functions within `group_by`. The disadvantage of this approach is that you need to specify all of the designed transformations upfront, otherwise richness of data will be gone, once the data is collapsed into the final summary statistics.

```{r}
## conventional summary
gapminder_plus %>% group_by(continent) %>% 
  summarise(min_lifeExp=min(lifeExp))

## two summary functions at the same time
gapminder_plus %>% group_by(continent) %>% 
  summarise_each(funs(min, max), lifeExp)

```

How can we delay the "collapsing" of the rich yearly data into the summary statistics without cluttering up the data frame? How can we summarize the data that includes more complex objects than single values (models, graphs, etc)?

# Using `dplyr`, `ggplot2` and `purrr` to explore a data set

It is intriguing to investigate data sets like `gapminder` for patterns. In this case , we're going to focus on just three variables to answer the question "How does life expectancy (`lifeExp`) change over time (`year`) for each country (`country`)?". A good place to start is with a plot:
_
```{r}

gapminder_plus %>% 
  ggplot() + 
  geom_line(mapping = aes(x = year, y = lifeExp, group = country, color = continent))
```

There seems to be an overall linear trend. Let's have a more detailed look, dividing the data set by continent:

```{r}
gapminder %>% 
  ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, color = country)) +
    geom_smooth(mapping = aes(x = year, y = lifeExp), method = "lm", color = "black") +
    facet_wrap( ~ continent) +
    theme(legend.position = "none")
```

There seems to be a good handful of countries with a pronounced irregularity. We will now produce a linear model for each country. But instead of repeating an action for each variable, we want to repeat an action for each country, a group of rows. Using "tidyverse" family of packages, it is possible to package the data belonging to each country and place it into a separate column as a "nested" data frame:

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()
by_country
```

What can we notice from this format. First of all our data frame have shrunk to only 142 observations - one for each country. The identifiers (country and continent) are still present in the dataset. There's a new data variable - by the type "tibble". it is possible to extract the nested elements of this data structure.

```{r}
by_country[["data"]][1] 
```


Now that we have our nested data frame, we're in a good position to fit some models. We will use a linear model function which we want to apply it to every data frame. 

# Introducing `map`

The function we are going to introduce next is going to allow us to repeat the same function for different subsets of our data. The same result can be accomplished with writing a loop, but the function we will consider next hides all of the "housekeeping" details (related to iterators and allocation of the output vector) and leaves only essential elements for the user to specify.

```{r, eval=FALSE}
`map(<DATA>, <FUNCTION>)'
```

What the function requires is the list (or vector) of data and a name of the function you want to apply to each element of your data.

> __NOTE:__  Simplest behaviour can be observed by calling a `map` function on a vector. Vector will be automatically transformed to a list and operation will be performed on each element of this list (i.e. on each element of the original vector). 
>
> If you specify the dataframe in place of `<DATA>`, because dataframe is a list of columns, the function will be performed on columns.
>
> If you want the function to operate in the slices of your dataframe (think of `group by` behavior) you will have to split the dataframe into the list of smaller dataframes, with each element containing only the data related to one slice.

The output of the `map` function is defined by the type of function used. There are special functions for each type of expected output: `map_chr()`, `map_dbl()`, `map_lgl()`, etc. The default function `map()` always outputs a list, so it should have been probably called `map_list()`. For two outputs you have a special version, which expects two data sets as input: `map2()`, etc. 

```{r}

# function sqrt applied to each element of the vector, output expected as a numeric vector
1:3 %>% map_dbl(sqrt)
# list of columns in, list of column means out
gapminder %>% select(year, gdpPercap, lifeExp) %>%
  map(mean)

# vector in, new variable out
gapminder %>% 
  mutate(gdpBillions = map2_dbl(gdpPercap , pop, function(x, y) x*y/10^9))
  
```


As you have seen above, we can use `map()` to apply a linear model to element of our complex `data` column. We will compute the results in a newly created variable in the `by_country` data frame with `mutate()`.

```{r}
by_country <- by_country %>% 
  mutate(model = purrr::map(data, ~lm(formula = lifeExp ~ year, data = .x)))
by_country
```

We have now created a linear model for every country! To take a better look at the model quality metrics, *e.g.* the R^2^, we will use the function `glance()` from the `broom` package, which is already loaded as part of `tidyverse`. If we apply it to a model, we get a data frame with a single row. We can use `mutate()` and `unnest()` to create a data frame with a row for each country. To suppress list columns (for data frames and models), we use `.drop = TRUE`.

```{r}
glance_df <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE)
glance_df
```

With this data frame in hand, we can start to look for models that don’t fit well:

```{r}
glance_df %>% 
  arrange(r.squared)
```

The worst models all appear to be in Africa. Let's double check that with a plot. Here we have a relatively small number of observations and a discrete variable, so `geom_jitter()` is effective:

```{r}
glance_df %>% 
  ggplot() + 
    geom_jitter(mapping = aes(x = continent, y = r.squared, color = continent), width = 0.3)
```

We could pull out the countries with particularly bad R^2^ and plot the data:

```{r}
bad_fit <- glance_df %>% 
  filter(r.squared < 0.5)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot() +
    geom_line(mapping = aes(x = year, y = lifeExp, colour = country))
```

We see two main effects here: the tragedies of the HIV/AIDS epidemic and the Rwandan genocide.
