---
title: "R for Reproducible Scientific Analysis"
subtitle: "Data Manipulation with dplyr"
author: "15 Feb 2018"
output:
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library("tidyverse")
gapminder <- gapminder::gapminder
```

# Dataframe Manipulation with `dplyr`

Manipulation of data is important to many researchers. We often select certain observations (rows) or variables (columns), we often group the data by (a) certain variable(s), and calculate summary statistics. Subsetting data sets can be challenging in base R, because there is a fair bit of repetition. Repeating yourself will cost you time, both now and later, and potentially introduce some nasty (silent) errors.

# Key Points

* Use the `dplyr` package to manipulate data
* Use `select()` to choose variables from data
* Use `filter()` to choose data based on values
* Use `group_by()` and `summarize()` to work with subsets of data
* Use `mutate()` to create new variables.

# The `dplyr` package

The [`dplyr`](https://cran.r-project.org/web/packages/dplyr/dplyr.pdf) package provides a number of useful functions for manipulating data in a way that will reduce  repetition, reduce the probability of making errors, and probably even save some typing. As an added bonus, the `dplyr` grammar is easier to read.

`dplyr` is part of the `tidyverse` package collection; it should already be loaded. If you've started a new session, load the `tidyverse` package:

```{r}
library("tidyverse")
```

# The pipe `%>%`

We often want to string together a series of `dplyr` functions. This is achieved using `dplyr`'s pipe operator, `%>%`. This takes the value on the left, and passes it as the first argument to the function call on the right. `%>%` isn’t limited to `dplyr` functions. It's an alternative way of writing any `R` code.

The shortcut to insert the pipe operator is `Ctrl`+`Shift`+`M` for Windows/Linux, and `Cmd`+`Shift`+`M` for Mac.

# Using `select()`

If, for example, we wanted to move forward with only a few of the variables of our data, we could use the `select()` function. This will keep only the variables you select.

```{r}
gapminder %>% select(year, country, gdpPercap)
year_country_gdp <- gapminder %>% 
  select(year, country, gdpPercap)
year_country_gdp
```

To help you understand why we wrote that in that way, let's walk through it step by step. First we summon the `gapminder` dataframe and pass it on, using the pipe symbol (`%>%`), to the next step, which is the `select()` function. In this case we don't specify which data object we use in the `select()` function since in gets that from the previous pipe. Fun fact: There is a good chance you have encountered pipes before in the shell. In `R`, a pipe symbol is `%>%` while in the shell it is `|` but the concept is the same!

# Using `filter()`

The `filter` function is a intuitive way of subsetting data. For example, we can look at the population per continent in 2002:

```{r}
gapminder %>% 
  filter(year == 2002) %>% 
  ggplot(mapping = aes(x = continent, y = pop)) +
  geom_boxplot()
```

If we now wanted to move forward with the above, but only with European countries, we can combine `select()` and `filter()`:

```{r}
year_country_gdp_euro <- gapminder %>% 
  filter(continent == "Europe") %>%
  select(year, country, gdpPercap)
year_country_gdp_euro
```

## <span style="color:red">Challenge 1</span>

Write a single command (which can span multiple lines and includes pipes) that will produce a dataframe that has the Norwegian values for `gdpPercap`, `lifeExp` and `year`, but not for other countries How many rows does your dataframe have and why?

```{r}
country_lifeExp_Norway <- gapminder %>%
  filter(country == "Norway") %>%
  select(year, lifeExp, gdpPercap)
country_lifeExp_Norway
```

As with last time, first we pass the gapminder dataframe to the `filter()` function, then we pass the filtered version of the gapminder dataframe to the `select()` function. **Note**: The order of operations is very important in this case. If we used `select()` first, `filter()` would not be able to find the variable continent since we would have removed it in the previous step.

# Using `group_by()`

Now, we were supposed to be reducing the error prone repetitiveness of what can be done with base `R`, but up to now we haven't done that since we would have to repeat the above for each continent. Instead of `filter()`, which will only pass observations that meet your criteria (in the above: `continent == "Europe"`), we can use `group_by()`, which will essentially use every unique criteria that you could have used in filter.

```{r}
gapminder
gapminder %>% 
  group_by(continent)
```

# Using `summarize()`

The above was a bit on the uneventful side because `group_by()` is much more exciting in conjunction with `summarize()`. This will allow use to create new variables by using functions that repeat for each of the continent-specific data frames. That is to say, using the `group_by()` function, we split our original dataframe into multiple pieces, then we can run functions (e.g. `mean()` or `sd()`) within `summarize()`).

```{r}
gapminder %>% 
  group_by(continent) %>%
  summarize(mean_gdpPercap = mean(gdpPercap))
```

We can also use a pipe to directly "feed" our processed data into `ggplot2`:

```{r}
gapminder %>% 
  group_by(continent) %>%
  summarize(mean_gdpPercap = mean(gdpPercap)) %>% 
  ggplot(mapping = aes(x = continent, y = mean_gdpPercap)) +
  geom_point()
```

That allowed us to calculate the mean `gdpPercap` for each continent, but it gets even better.

## <span style="color:red">Challenge 2</span>

Calculate the average life expectancy per country in Asia. Which has the longest average life expectancy and which has the shortest average life expectancy?

```{r}
gapminder %>% 
  filter(continent == "Asia") %>% 
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>% 
  filter(mean_lifeExp == min(mean_lifeExp) | mean_lifeExp == max(mean_lifeExp))
```

The function `group_by()` allows us to group by multiple variables. Let's group by year and continent.

```{r}
gapminder %>% 
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap))
```

That is already quite powerful, but it gets even better! You're not limited to defining one new variable in `summarize()`.

```{r}
gapminder %>% 
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap),
            sd_gdpPercap = sd(gdpPercap),
            mean_pop = mean(pop),
            sd_pop = sd(pop))
```

# Using `mutate()`

We can also create new variables prior to (or even after) summarizing information using `mutate()`.

```{r}
gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9)

gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9) %>%
  group_by(continent, year) %>%
  summarize(mean_gdpPercap = mean(gdpPercap),
            sd_gdpPercap = sd(gdpPercap),
            mean_pop = mean(pop),
            sd_pop = sd(pop),
            mean_gdp_billion = mean(gdp_billion),
            sd_gdp_billion = sd(gdp_billion))
```

## <span style="color:red">Challenge 3</span>

Calculate the average life expectancy and GDP (in billions) in 1987 for each continent.

```{r}
gapminder %>% 
  mutate(gdp_billion = gdpPercap * pop / 10^9) %>%
  filter(year == 1987) %>%
  group_by(continent) %>%
  summarize(mean_lifeExp = mean(lifeExp),
            mean_gdp_billion = mean(gdp_billion))
```

# Using `dplyr` and `ggplot2` to visualize average global life expectancy

We will now demonstrate how to use functions from both `dplyr`and `ggplot2` to visualize the average life expectancy of countries that are part of the `gapminder` data set.

First, we need to calculate the average life expectancy for each country using the `summarize()` function:

```{r}
gapminder_country_summary <- gapminder %>% 
  group_by(country) %>%
  summarize(mean_lifeExp = mean(lifeExp))
gapminder_country_summary
```

Next, we will re-use a map introduced in `ggplot` lesson to align with the average life expectancy data from above. We can merge this data with out own, by ensuring the map data has a _common variables_ as our statistics data file from gapminder. In this case, our world map has a column called `NAME` that corresponds with a column in the gapminder data called `country`, so we just indicate this in the `by` argument and merge the two data sets together. We can then plot different aesthetics on the map, based on the data from gampinder.

```{r}
# install.packages("rnaturalearth")
world_map <- rnaturalearth::ne_countries(returnclass = "sf")
gapminder_map <- right_join(gapminder_country_summary, world_map, by=c("country"="name"))

ggplot(gapminder_map) +
  geom_sf(aes(fill = mean_lifeExp))+
  scale_fill_viridis_c()+
  theme_void()+
  coord_sf()
```

Note that "grey" countries are not part of `gapminder` or appear under a different name and hence only appear as `NA`'s.


# Key Points

* Use the `dplyr` package to manipulate data
* Use `select()` to choose variables from data
* Use `filter()` to choose data based on values
* Use `group_by()` and `summarize()` to work with subsets of data
* Use `mutate()` to create new variables
